Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-02-28 agujournaltemplate.tex
Focus Phase Day 12 - Intro almost done, start Discussion

    % Phytoplankton is important
    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of photoautotrophic microorganisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.

    - Phytoplankton communities can be described from taxonomic/functional group aspects, but also from biodiversity aspects (or trait-based, but this I am not doing here)

@@

    The loss of biodiversity is predicted to have a lasting impact on the functioning of an ecosystem, since links have been found between the richness of a phytoplankton community and the biomass and resource use efficiency \cite{ptacnik_diversity_2008, lehtinen_phytoplankton_2017}.

    - mention BDEF in Tropics paper, or some other paper, to highlight that the tropics are notoriously understudied in relation to phytoplankton diversity.

    %%% This paper is clearly not BDEF, but environmental impacts on diversity and community!section

@@

    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Describe CARIACO changes in ecosystem

    %% Changes in Functional Groups

@@

    %% in depth look at biodiversity aspect, in particular microscopy data

@@

    - AMO (ref to 2019 Review)
    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al. 2015). These climatic changes had a profound impact throughout the Atlantic (Alexander et al. 2014, Nye et al. 2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.

    - Finally, says in Review about Phytoplankton community
    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

@@

\section{Discussion}
This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem, that is driven by local environmental variability, which is in turn affected by large scale climatic shifts. In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime and ...

- The 20 year record of CARIACO time series is a great opportunity to better understand the connection between pelagic proccesses, climate and sedimentation.

- Our main results show that...

>>>>

    % Phytoplankton is important
    Phytoplankton provides the foundation for marine food webs and plays a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly determined by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
    %%% Mention Top-Down and Bottom-Up forces structuring communities and affection biodiversity.

@@

    The loss of biodiversity is predicted to have a lasting impact on the functioning of an ecosystem, since links have been found between the richness of a phytoplankton community and ecosystem functions, such as biomass and resource use efficiency \cite{ptacnik_diversity_2008, lehtinen_phytoplankton_2017}. Such links need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014} and under greater threat of extinction, when compared to temperate regions \cite{finnegan_paleontological_2015}.
    % DON'T FORGET TO MENTION CLIMATE IMPACTS (as I investigate AMO linkage and ENSO MEI etc.)
    % this would actually make the most sense here!

@@

    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    %%% mention climate impacts (AMO + ENSO) here!

    % Describe general seasonal cycle with functional groups etc.
    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids and nanoplankton also contributing. In the low-wind rainy season, the Cariaco basin exhibits oligotrophic conditions, where the portion of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % Describe CARIACO changes in ecosystem
    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
    The trends in microscopy community data were confirmed and expanded by \citeA{pinckney_phytoplankton_2015}. They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC). This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}. In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition. These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.

    % NOw we come in!
    - The question arises, what were the drivers of change in phytoplankton community?
    - Bottom-up or top-down processes? response to regional changes in oceanographic conditions, in turn driven by climate change or oscillations?
    - Highlight, that there is a large mismatch in data-coverage particularly for Top-Down aspects, fisheries is landing data (hard to quantify anthropogenic impact on this data and ecosystem) and Zooplankton data is only from after 2002. Also HPLC data is lacking in coverage.
    - We focus on Bottom-up, given the available data, and try to identify changes in biodiversity and community and relate these to both local effects and climatic drivers.

    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    % "Changes in plankton biodiversity in response to regional changes in oceangraphic conditions" driven by climate oscillaitons (from REVIEW 2019).
    %% Changes in Functional Groups

@@

    % FOCUS ON Community Changes + BOTTOM UP

    These changes are believed to be influenced by various factors, including climate variability, nutrient fluxes, and ocean circulation patterns.
    % WHAT ARE THE DRIVERS OF COMMUNITY CHANGE? Bottom up OR top down, we focus on bottom up...

    Describe physical changes!

    - AMO (ref to 2019 Review)
    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al. 2015). These climatic changes had a profound impact throughout the Atlantic (Alexander et al. 2014, Nye et al. 2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
    - ENSO teleconnection (ref to 2019 Review)
    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).

     - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

@@

    - Finally, add the info from Review about Phytoplankton community, about community and upwelling returning in 2014 to previous levels/state.
    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

@@

\section{Discussion}
This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.

- Our main results show that...
1. we see Overlap in yearly mean anomalies between certain variables

2. mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift

3. Clustering shows return of community to previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly goes along gradient separating the two cluster (NMDS2)

3.1 Turnover is different between two clusters, also corresponding to evennness between clusters

3.2 Discuss which variables actually vary between clusters and how they could mechanistically explain difference between communities

4. Gradient Forest clearly identifies climate indices as best predictors of community change, then nutrients etc.
This is really interesting, as it would be expected, that in situ-measurements are best predictor of community. BUT either the community + top down processes are capturing more of that variance, or we simply are missing a bottom-up variable that can explain it.
but that most likely requires mechanistic modeling to investigate further...

5. Compare to what happened at other long term time series in the ocean...
